Towards a Practical Face Recognition System: \\ Robust Alignment and Illumination by Sparse Representation

### 2011-04-04 faces_system.tex
Vspacing from zihan

%\documentclass[12pt,journal,draftcls,letterpaper,onecolumn]{IEEEtran}
%\documentclass[9.5pt,journal,final,finalsubmission,twocolumn]{IEEEtran}

@@

\usepackage{times}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amscd}
\usepackage{amssymb}

%\usepackage{subfigure}

%\ifCLASSOPTIONcompsoc
 %\usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
%\else
 %\usepackage[caption=false,font=footnotesize]{subfig}
%\fi
\usepackage[caption=false,font=footnotesize]{subfig}

\usepackage{pdfsync}
\usepackage{helvet}
\usepackage{url}
\usepackage{color}

@@

@@

\begin{figure} \centering \begin{tabular}{cc}
\includegraphics[height=1in]{figures_cvpr/promo/case1/detector.png}&
\hspace{3mm}
\includegraphics[height=1in]{figures_cvpr/promo/case1/sci_with_axis_face_case1.png}
\\ \includegraphics[height=1in]{figures_cvpr/promo/alignment_and_detector.png}&
\hspace{3mm}
\includegraphics[height=1in]{figures_cvpr/promo/case2/sci_with_axis_face_case2.png}
\\ \includegraphics[height=1in]{figures_cvpr/promo/case3/alignment.png} &
\hspace{3mm}
\includegraphics[height=1in]{figures_cvpr/promo/case3/sci_with_axis_face_case3.png}
\end{tabular} \caption{{\bf Effects of registration and illumination on
Recognition}. In this example we identify the girl among 20 subjects, by
computing the sparse representation of her input face with respect to the
entire training set. The absolute sum of the coefficients associated with each
subject is plotted on the right. We also show the faces reconstructed with each
subject's training images weighted by the associated sparse coefficients. The
red line (cross) corresponds to her true identity, subject 12. {\bf Top:} The
input face is from Viola and Jones' face detector (the black box) and all 38
illuminations specified in Section \ref{sec:illumination} are used in the
training.  {\bf Middle:} The input face is well-aligned (the white box) with
the training by our algorithm specified in Section \ref{sec:registration} but
only 24 frontal illuminations are used in the training for recognition (see
Section \ref{sec:illumination}). {\bf Bottom:} The input face is well aligned and
a sufficient set (all 38) of
illuminations are used in the training. Both are necessary for correct recognition
using SRC.} \label{fig:promo}
\vspace{0mm}
\end{figure}
%\subsection{Compound Effect of Registration and Illumination}

@@

\subsection{Related Work}

We briefly review existing techniques for recognition, image registration, and
handling of illumination variation.  Our system is based purely on 2D
techniques.  This fact immediately distinguishes our approach from systems
that either require a 3D data as an input, or attempt to estimate a 3D model
from 2D input \cite{amberg2007reconstructing,Blanz2003-PAMI}.  While these
techniques can achieve better robustness to pose variation given
a sufficiently accurate 3D model, for access control applications where only
moderate pose variation is present, the proposed method will be more
than sufficient.   Note that 2D images of faces
under varying illuminations already contain 3D shape-related information,\footnote{In principle,
one can recover the 3D shape of the face from multiple illuminations using photometric stereo.}
and this information can be leveraged by 2D algorithms for alignment and recognition
even if shape is not reconstructed explicitly.
%In addition, the proposed sparsity-based robust
%alignment and recognition techniques are not limited to 2D data and can be easily extended
%to 3D data if necessary.

@@

% Illumination
Researchers have tried various techniques to deal with illumination variation.
In almost all recognition algorithms where only a single gallery image is
available per individual, illumination effects are regarded as a nuisance that
must be removed before the algorithm can continue.  This is typically done by
making statistical assumptions about how illumination affects the image, and
using those assumptions to extract a new representation that is claimed to be
illumination invariant.  Recent examples include \cite{chen2006total} and
\cite{zhou2007appearance}.  However, despite these efforts, truly
illumination-invariant features are difficult to obtain from a single input
image.  We argue that if one has the luxury of designing the acquisition system
and the application demands a high recognition rate,
it is then unwise to limit the gallery to a
single image per person.  We therefore take the strategy of sampling many
gallery images of each individual under varying illuminations.  These images
are used as the basis for either a convex cone model
\cite{Georghiades2001-PAMI,belhumeur1998set}, or a subspace model
\cite{Basri2003-PAMI}.  Images are captured using a simple-to-construct
projector based light stage.  While similar systems have been used for
other applications, to our knowledge, we
are the first to use projectors to indirectly illuminate a subject's face for
the purpose of face recognition.

\subsection{Contributions} In this paper, we show how
registration and illumination
can be simultaneously addressed within a robust sparse representation
framework. We show that face registration, a challenging
nonlinear problem, can be solved by a series of linear programs
that iteratively minimize the sparsity of the registration
error. This leads to an efficient and effective alignment
algorithm for face images that works for a large range of
variation in translation, rotation, and scale, even when
the face is only partially visible due to eyeglasses, closed
eyes and open mouth, sensor saturation, etc.  We also propose a
sufficient set of training illuminations
that is capable of linearly representing typical indoor and outdoor
lighting, along with a practical hardware system for capturing
them.

@@

We have gathered the statistics of the Viola and Jones'
 face detector on the Multi-PIE dataset. For 4,600 frontal
 images of 230 subjects under 20 different illuminations,
 using manual registration as the ground truth, the average
 misalignment error of the detected faces is about 6 pixels
 and the average variation in scale is 8\%. This falls
 safely inside the region of attraction for our alignment
 algorithm.
\begin{figure*}
\centering
{\small
\begin{tabular}{ccc}
\includegraphics[height=1.4in]{figures_pami/x_y_roa.png} &
\includegraphics[height=1.4in]{figures_pami/y_theta_roa.png} &
\includegraphics[height=1.4in]{figures_pami/y_s_roa.png}\\
(a)&(b)&(c)
\end{tabular}
\begin{tabular}{cccc}
\includegraphics[height=1in]{figures_pami/x_tr.png} &
\includegraphics[height=1in]{figures_pami/y_tr.png} &
\includegraphics[height=1in]{figures_pami/theta.png} &
\includegraphics[height=1in]{figures_pami/scale.png}\\
(d)&(e)&(f)& (g)
\end{tabular}
\vspace{0mm}
}
\caption{{\bf Region of attraction.} Fraction of subjects for which the algorithm successfully aligns a synthetically perturbed test image.  The amount of translation is expressed as a fraction of the distance between the outer eye corners, and the amount of in-plane rotation in degrees. {\bf Top row:} (a) Simultaneous translation in $x$ and $y$ directions. (b) Simultaneous translation in $y$ direction and in-plane rotation. (c) Simultaneous translation in $y$ direction and scale variation. {\bf Bottom row:} (d) Translation in $x$ direction only. (e) Translation in $y$ direction only. (f) In-plane rotation only. (g) Scale variation only.}
\label{fig:attraction}
\vspace{0mm}
\end{figure*}

\noindent{2) \em Multiscale Implementation.}
Performing alignment in a multiscale fashion has two benefits: first, it provides a larger region of attraction, and second, it reduces overall computational cost. Here, we further investigate the convergence behavior of the algorithm as a function of the standard deviation $\sigma$ of the Gaussian smoothing filter and the number of scales considered.
We use the same 7 illuminations in
Session 1 as training, and all 20 illuminations in the same
session as testing. We introduce artificial deformation in
both $x$ and $y$ directions up to 16 pixels in the
$80\times 60$ frame, with a step size of 4 pixels, i.e.,
$(\Delta x, \Delta y) \in \{-16,-12,\ldots,12,16\} \times
\{-16,-12,\ldots,12,16\}$. We consider an alignment
successful if the estimated coordinates of the eye-corners
are within 1 pixel from the ground truth in the original
image.  In Figure \ref{fig:multiscale}, we report the
alignment success rate, averaged over the artificially
perturbed initial deformations, as a function of the
standard deviation of the Gaussian kernel $\sigma$, for
three choices of the number of scales. As one can see,
using multiscale indeed improves the performance, and when
3 scales are used, a smaller convolution kernel can achieve
a similar performance compared to a much larger kernel when
only 2 scales are used.
\begin{figure}
\centering
\includegraphics[width=3.5in]{figures_pami/multiscale.png}
\vspace{0mm}
\caption{{\bf Multiscale alignment.} This figure shows the average success rate of alignment over all possible perturbations. A smaller blur kernel can be applied to achieve certain level of performance when more scales are used.}
\label{fig:multiscale}
\vspace{0mm}
\end{figure}

\noindent{3) \em 3D Pose Variation.} As densely sampled pose and
    illumination face images are not available in any of
    the public databases, including Multi-PIE, we have
    collected our own dataset using our own system (to be
    introduced in the next section). We use frontal face
    images of a subject under the 38 illuminations proposed
    in the next section as training. For testing, we
    collect images of the subject under a typical indoor
    lighting condition at pose ranging from $-90^\circ$ to
    $+90^\circ$ with step size 5.625$^\circ$, a total of 33
    poses. We use Viola and Jones' face detector to
    initialize our alignment algorithm.
Figure \ref{fig:pose-alignment} shows that our algorithm works reasonably well with poses up
to $\pm 45^\circ$.
Note that this level of out-of-plane
 pose variation is beyond what we intend to handle with our formulation.
\begin{figure}
\centering
{\small
\begin{tabular}{ccccc}
\includegraphics[height=0.55in]{figures_cvpr/5} &
\includegraphics[height=0.55in]{figures_cvpr/7} &
\includegraphics[height=0.55in]{figures_cvpr/09} &
\includegraphics[height=0.55in]{figures_cvpr/11} &
\includegraphics[height=0.55in]{figures_cvpr/13} \vspace{0mm} \\ 
(a) & (b) & (c) & (d) & (e)\\
\includegraphics[height=0.55in]{figures_cvpr/15} &
\includegraphics[height=0.55in]{figures_cvpr/17} &
\includegraphics[height=0.55in]{figures_cvpr/19} &
\includegraphics[height=0.55in]{figures_cvpr/21} &
\includegraphics[height=0.55in]{figures_cvpr/3} \vspace{0mm} \\ 
(f) & (g) & (h) & (i) & (j) \vspace{0mm}
\end{tabular}
}
\caption{{\bf 2D Alignment of test images with different poses to frontal training images.} {\bf (a) to (i):}  plausible alignment for pose from $-45^{\circ}$
to $+45^{\circ}$. {\bf (j):} a case when the algorithm fails for an extreme pose ($>45^{\circ}$).
}\label{fig:pose-alignment} \vspace{0mm}
\end{figure}

\subsection{Comparison with Related Work}
Our modification to SRC roots solidly in the tradition of
adding deformation-robustness to face recognition algorithms
\cite{Cootes2001-PAMI,Gross2006-PAMI,Wiskott1997-PAMI}.
However, the only previous work to investigate face alignment
in the context of sparse signal representation and SRC is the
work of \cite{Huang2008-CVPR}. They consider the case where the
training images themselves are misaligned and allow one
deformation per training image. They linearize the training
rather than the test, which is computationally more costly as
it effectively triples the size of the training set. In
addition, as they align the test image to all subjects
simultaneously, it potentially is more prone to local minima
when the number of subjects increases, as we will see in the
following experimental comparisons.
\begin{enumerate}
\item {\em Extended Yale B.} In this experiment, we have
    used the same experimental settings as in
    \cite{Huang2008-CVPR}. 20 subjects are selected and
    each has 32 frontal images (selected at random) as
    training and another 32 for testing. An artificial
    translation of 10 pixels (in both $x$ and $y$
    directions) is introduced to the test image. For our
    algorithm we downsample all the images to $88\times
    80$ for memory reasons, whereas the work of
    \cite{Huang2008-CVPR} uses random projections.
Note that the use of cropped images in this experiment introduces image boundary effects.
    Our
    algorithm achieves the recognition rate 93.7\%,
    compared to 89.1\% recognition rate reported in
    \cite{Huang2008-CVPR}.
\item {\em CMU Multi-PIE.} In this experiment, we choose
    all subjects from the CMU Multi-PIE database, 7
    training images from Session 1 and 1 test image from
    Session 2 per person. The setting is exactly the same
    as the previous experiment on 2D deformation. We again
    work with downsampled images of size $80\times 60$ pixels. An
    artificial translation of 5 pixels (in both $x$ and $y$
    directions) was induced in the test image. The
    algorithm of \cite{Huang2008-CVPR} achieves a
    recognition rate of 67.5\%,\footnote{That algorithm has
    two free parameters - $l$ and $d$, which govern the tradeoff between
	accuracy and run-time. For this experiment
    we chose $l = 1$ and $d = 514$.} while ours achieves 92.2\%.
    \end{enumerate}

@@

We have designed a system that can acquire frontal images of a subject while
simultaneously illuminating the subject from all directions above horizontal. A sketch of the
system is shown in Figure \ref{fig:system}: The illumination
system consists of four projectors that display various bright
patterns onto the three white walls in the corner of a dark
room.  The light reflects off of the walls and illuminates the
user's head indirectly.  After taking the frontal illuminations
we rotate the chair by 180 degrees and take pictures from the
opposite direction.  Having two cameras speeds the process
since only the chair needs to be moved in between frontal and
rear illuminations. Our projector-based system has several
advantages over flash-based illumination systems for face recognition:
\begin{itemize}
\item The illuminations can be modified in software, rather than hardware.
\item It is easy to capture many different illuminations quickly.
\item Good coverage and distant illumination can be achieved simultaneously.
\item There is no need to mount anything on the walls or construct a large dome.
\item The system can be assembled from off-the-shelf hardware.
\end{itemize}
\begin{figure}
\centerline{\hspace{-0.1in}\includegraphics[height=2.3in]{figures_cvpr/camera_rig.pdf}\vspace{0mm}}
\caption{{\bf Training acquisition system:} Four projectors and two cameras controlled by one computer.}
\vspace{-2em}
\label{fig:system}
\end{figure}
With our projector system, our choice of illuminations is
constrained only by the need to achieve a good
SNR\footnote{Since illuminations with more pixels illuminated
will have a better SNR (provided they don't saturate), there is
an engineering tradeoff between the SNR and the number of
training images.}, avoid saturation, and achieve
a reasonably short acquisition time.  Two simplifying
assumptions that we make are that every pixel is either turned
fully on or off in every illumination, and that the illuminated
regions do not overlap.

@@

\begin{figure}
\centering
\begin{tabular}{@{}c@{}c@{}}
\includegraphics[height=1.35in]{figures_cvpr/illum_results/coverage_sunset.pdf} &
\includegraphics[height=1.35in]{figures_cvpr/illum_results/granularity_sunset.pdf} \\
(a) Coverage & (b) Granularity
\end{tabular}\vspace{0mm}
\caption{{\bf Study of sufficient illuminations.} The average $\ell^1$ registration residual versus different illumination training sets. }
\label{fig:illumination-sufficiency}
\vspace{-4.5em}
\end{figure}

@@

See below for the 38 training images of one subject:
\begin{figure}[h]
\centering
\includegraphics[width=3.5in]{figures_cvpr/training.png}
\vspace{-2em}
\end{figure}

@@

\begin{table}
\caption{Recognition rates on the Multi-PIE database for
Algorithm 1 and \cite{Yang2010-CVPR}}
\centerline{
\begin{tabular}{|c|c|c|c|c| }
\hline
Recognition rate & Session 2 & Session 3 & Session 4  \\
\hline
{Algorithm 1, $S=1$} & 90.7\% & 89.6\% & 87.5\% \\
\hline
{Algorithm 1} & 93.9\% & 93.8\% & 92.3\% \\
\hline
{Algorithm 1 with improved window} & 95.0\% & {\bf 96.3}\% & {\bf 97.3}\% \\
\hline
\cite{Yang2010-CVPR} & {\bf 95.2}\% & 93.4\% & 95.1\% \\
\hline
\end{tabular}
\label{tab:MultiPIE-recognition2} }
\end{table}

\newcommand{\tempwidth}[0]{0.58in}
\begin{figure}
\centering
{\small
\begin{tabular}{@{}c@{}c@{}c@{}c@{}c@{}c@{}}
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/079_01_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/111_01_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/196_01_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/130_01_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/163_01_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/175_01_01_051_08.png} \\
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/079_02_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/111_02_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/196_02_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/130_02_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/163_02_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/175_02_01_051_08.png} \\
(a) & (b) & (c) & (d) & (e) & (f) \vspace{0mm}
\end{tabular}
}
\caption{{\bf Representative failures from Multi-PIE}. {\bf Top:} training from Session 1; {\bf Bottom:} test images from Session 2. Due to changes in hair, glasses, beard, or pose, our alignment fails on these subjects regardless of test image illumination.}
\label{fig:failed-examples}
\vspace{-15mm}
\end{figure}

>>>>

%\documentclass[12pt,journal,draftcls,letterpaper,onecolumn]{IEEEtran}
%\documentclass[9.5pt,journal,final,compsoc,twocolumn]{IEEEtran}

@@

\usepackage{times}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amscd}
\usepackage{amssymb}
\usepackage{subfigure}
\usepackage{pdfsync}
\usepackage{helvet}
\usepackage{url}
\usepackage{color}

@@

% \usepackage[small]{caption}

@@

\begin{figure}
\centering \begin{tabular}{cc}
\includegraphics[height=1in]{figures_cvpr/promo/case1/detector.png}&
\hspace{3mm}
\includegraphics[height=1in]{figures_cvpr/promo/case1/sci_with_axis_face_case1.png}
\\ \includegraphics[height=1in]{figures_cvpr/promo/alignment_and_detector.png}&
\hspace{3mm}
\includegraphics[height=1in]{figures_cvpr/promo/case2/sci_with_axis_face_case2.png}
\\ \includegraphics[height=1in]{figures_cvpr/promo/case3/alignment.png} &
\hspace{3mm}
\includegraphics[height=1in]{figures_cvpr/promo/case3/sci_with_axis_face_case3.png}
\end{tabular} \caption{{\bf Effects of registration and illumination on
Recognition}. In this example we identify the girl among 20 subjects, by
computing the sparse representation of her input face with respect to the
entire training set. The absolute sum of the coefficients associated with each
subject is plotted on the right. We also show the faces reconstructed with each
subject's training images weighted by the associated sparse coefficients. The
red line (cross) corresponds to her true identity, subject 12. {\bf Top:} The
input face is from Viola and Jones' face detector (the black box) and all 38
illuminations specified in Section \ref{sec:illumination} are used in the
training.  {\bf Middle:} The input face is well-aligned (the white box) with
the training by our algorithm specified in Section \ref{sec:registration} but
only 24 frontal illuminations are used in the training for recognition (see
Section \ref{sec:illumination}). {\bf Bottom:} The input face is well aligned and
a sufficient set (all 38) of
illuminations are used in the training. Both are necessary for correct recognition
using SRC.}\label{fig:promo}
\vspace{-4mm}
\end{figure}
%\subsection{Compound Effect of Registration and Illumination}

@@

\subsection{Related Work}
We briefly review existing techniques for recognition, image
registration, and handling of illumination variation.  Our
system is based purely on 2D techniques.  This fact immediately
distinguishes our approach from systems that either require a
3D data as an input, or attempt to estimate a 3D model from 2D
input \cite{amberg2007reconstructing,Blanz2003-PAMI}.  While
these techniques can achieve better robustness to pose
variation given a sufficiently accurate 3D model, for access
control applications where only moderate pose variation is
present, the proposed method will be more than sufficient.
Note that 2D images of faces under varying illuminations
already contain 3D shape-related information,\footnote{In
principle, one can recover the 3D shape of the face from
multiple illuminations using photometric stereo.} and this
information can be leveraged by 2D algorithms for alignment and
recognition even if shape is not reconstructed explicitly.
%In addition, the proposed sparsity-based robust
%alignment and recognition techniques are not limited to 2D data and can be easily extended
%to 3D data if necessary.

@@

% Illumination
Researchers have tried various techniques to deal with illumination variation.
In almost all recognition algorithms where only a single gallery image is
available per individual, illumination effects are regarded as a nuisance that
must be removed before the algorithm can continue.  This is typically done by
making statistical assumptions about how illumination affects the image, and
using those assumptions to extract a new representation that is claimed to be
illumination invariant.  Recent examples include \cite{chen2006total} and
\cite{zhou2007appearance}.  However, despite these efforts, truly
illumination-invariant features are difficult to obtain from a single input
image.  We argue that if one has the luxury of designing the acquisition system
and the application demands a high recognition rate,
it is then unwise to limit the gallery to a
single image per person.  We therefore take the strategy of sampling many
gallery images of each individual under varying illuminations.  These images
are used as the basis for either a convex cone model
\cite{Georghiades2001-PAMI,belhumeur1998set}, or a subspace model
\cite{Basri2003-PAMI}.  Images are captured using a simple-to-construct
projector based light stage.  While similar systems have been used for
other applications, to our knowledge, we
are the first to use projectors to indirectly illuminate a subject's face for
the purpose of face recognition.
\subsection{Contributions} In this paper, we show how
registration and illumination
can be simultaneously addressed within a robust sparse representation
framework. We show that face registration, a challenging
nonlinear problem, can be solved by a series of linear programs
that iteratively minimize the sparsity of the registration
error. This leads to an efficient and effective alignment
algorithm for face images that works for a large range of
variation in translation, rotation, and scale, even when
the face is only partially visible due to eyeglasses, closed
eyes and open mouth, sensor saturation, etc.  We also propose a
sufficient set of training illuminations
that is capable of linearly representing typical indoor and outdoor
lighting, along with a practical hardware system for capturing
them.

@@

We have gathered the statistics of the Viola and Jones'
 face detector on the Multi-PIE dataset. For 4,600 frontal
 images of 230 subjects under 20 different illuminations,
 using manual registration as the ground truth, the average
 misalignment error of the detected faces is about 6 pixels
 and the average variation in scale is 8\%. This falls
 safely inside the region of attraction for our alignment
 algorithm.
\begin{figure*}
\centering
{\small
\begin{tabular}{ccc}
\includegraphics[height=1.2in]{figures_pami/x_y_roa.png} &
\includegraphics[height=1.2in]{figures_pami/y_theta_roa.png} &
\includegraphics[height=1.2in]{figures_pami/y_s_roa.png}\\
(a)&(b)&(c)
\end{tabular}
\begin{tabular}{cccc}
\includegraphics[height=1in]{figures_pami/x_tr.png} &
\includegraphics[height=1in]{figures_pami/y_tr.png} &
\includegraphics[height=1in]{figures_pami/theta.png} &
\includegraphics[height=1in]{figures_pami/scale.png}\\
(d)&(e)&(f)& (g)
\end{tabular}
\vspace{0mm}
}
\caption{{\bf Region of attraction.} Fraction of subjects for which the algorithm successfully aligns a synthetically perturbed test image.  The amount of translation is expressed as a fraction of the distance between the outer eye corners, and the amount of in-plane rotation in degrees. {\bf Top row:} (a) Simultaneous translation in $x$ and $y$ directions. (b) Simultaneous translation in $y$ direction and in-plane rotation. (c) Simultaneous translation in $y$ direction and scale variation. {\bf Bottom row:} (d) Translation in $x$ direction only. (e) Translation in $y$ direction only. (f) In-plane rotation only. (g) Scale variation only.}
\label{fig:attraction}
\vspace{0mm}
\end{figure*}

\noindent{2) \em Multiscale Implementation.}
Performing alignment in a multiscale fashion has two benefits: first, it provides a larger region of attraction, and second, it reduces overall computational cost. Here, we further investigate the convergence behavior of the algorithm as a function of the standard deviation $\sigma$ of the Gaussian smoothing filter and the number of scales considered.
We use the same 7 illuminations in
Session 1 as training, and all 20 illuminations in the same
session as testing. We introduce artificial deformation in
both $x$ and $y$ directions up to 16 pixels in the
$80\times 60$ frame, with a step size of 4 pixels, i.e.,
$(\Delta x, \Delta y) \in \{-16,-12,\ldots,12,16\} \times
\{-16,-12,\ldots,12,16\}$. We consider an alignment
successful if the estimated coordinates of the eye-corners
are within 1 pixel from the ground truth in the original
image.  In Figure \ref{fig:multiscale}, we report the
alignment success rate, averaged over the artificially
perturbed initial deformations, as a function of the
standard deviation of the Gaussian kernel $\sigma$, for
three choices of the number of scales. As one can see,
using multiscale indeed improves the performance, and when
3 scales are used, a smaller convolution kernel can achieve
a similar performance compared to a much larger kernel when
only 2 scales are used.
\begin{figure}
\centering
\includegraphics[width=3in]{figures_pami/multiscale.png}
\vspace{-2mm}
\caption{{\bf Multiscale alignment.} This figure shows the average success rate of alignment over all possible perturbations. A smaller blur kernel can be applied to achieve certain level of performance when more scales are used.}
\label{fig:multiscale}
\vspace{-4mm}
\end{figure}

\noindent{3) \em 3D Pose Variation.} As densely sampled pose
and
    illumination face images are not available in any of
    the public databases, including Multi-PIE, we have
    collected our own dataset using our own system (to be
    introduced in the next section). We use frontal face
    images of a subject under the 38 illuminations proposed
    in the next section as training. For testing, we
    collect images of the subject under a typical indoor
    lighting condition at pose ranging from $-90^\circ$ to
    $+90^\circ$ with step size 5.625$^\circ$, a total of 33
    poses. We use Viola and Jones' face detector to
    initialize our alignment algorithm.
Figure \ref{fig:pose-alignment} shows that our algorithm works reasonably well with poses up
to $\pm 45^\circ$.
Note that this level of out-of-plane
 pose variation is beyond what we intend to handle with our formulation.
\begin{figure}
\centering
{\small
\begin{tabular}{ccccc}
\includegraphics[height=0.5in]{figures_cvpr/5} &
\includegraphics[height=0.5in]{figures_cvpr/7} &
\includegraphics[height=0.5in]{figures_cvpr/09} &
\includegraphics[height=0.5in]{figures_cvpr/11} &
\includegraphics[height=0.5in]{figures_cvpr/13} \vspace{0mm} \\
(a) & (b) & (c) & (d) & (e)\\
\includegraphics[height=0.5in]{figures_cvpr/15} &
\includegraphics[height=0.5in]{figures_cvpr/17} &
\includegraphics[height=0.5in]{figures_cvpr/19} &
\includegraphics[height=0.5in]{figures_cvpr/21} &
\includegraphics[height=0.5in]{figures_cvpr/3} \vspace{0mm} \\
(f) & (g) & (h) & (i) & (j) \vspace{0mm}
\end{tabular}
}
\caption{{\bf 2D Alignment of test images with different poses to frontal training images.} {\bf (a) to (i):}  plausible alignment for pose from $-45^{\circ}$
to $+45^{\circ}$. {\bf (j):} a case when the algorithm fails for an extreme pose ($>45^{\circ}$).
}\label{fig:pose-alignment} \vspace{-4mm}
\end{figure}
\subsection{Comparison with Related Work}
Our modification to SRC roots solidly in the tradition of
adding deformation-robustness to face recognition algorithms
\cite{Cootes2001-PAMI,Gross2006-PAMI,Wiskott1997-PAMI}.
However, the only previous work to investigate face alignment
in the context of sparse signal representation and SRC is the
work of \cite{Huang2008-CVPR}. They consider the case where the
training images themselves are misaligned and allow one
deformation per training image. They linearize the training
rather than the test, which is computationally more costly as
it effectively triples the size of the training set. In
addition, as they align the test image to all subjects
simultaneously, it potentially is more prone to local minima
when the number of subjects increases, as we will see in the
following experimental comparisons.
\begin{enumerate}
\item {\em Extended Yale B.} In this experiment, we have
    used the same experimental settings as in
    \cite{Huang2008-CVPR}. 20 subjects are selected and
    each has 32 frontal images (selected at random) as
    training and another 32 for testing. An artificial
    translation of 10 pixels (in both $x$ and $y$
    directions) is introduced to the test image. For our
    algorithm we downsample all the images to $88\times
    80$ for memory reasons, whereas the work of
    \cite{Huang2008-CVPR} uses random projections.
Note that the use of cropped images in this experiment introduces image boundary effects.
    Our
    algorithm achieves the recognition rate 93.7\%,
    compared to 89.1\% recognition rate reported in
    \cite{Huang2008-CVPR}.
\item {\em CMU Multi-PIE.} In this experiment, we choose
    all subjects from the CMU Multi-PIE database, 7
    training images from Session 1 and 1 test image from
    Session 2 per person. The setting is exactly the same
    as the previous experiment on 2D deformation. We again
    work with downsampled images of size $80\times 60$ pixels. An
    artificial translation of 5 pixels (in both $x$ and $y$
    directions) was induced in the test image. The
    algorithm of \cite{Huang2008-CVPR} achieves a
    recognition rate of 67.5\%,\footnote{That algorithm has
    two free parameters - $l$ and $d$, which govern the tradeoff between
	accuracy and run-time. For this experiment
    we chose $l = 1$ and $d = 514$.} while ours achieves 92.2\%.
    \end{enumerate}

@@

We have designed a system that can acquire frontal images of a subject while
simultaneously illuminating the subject from all directions above horizontal. A sketch of the
system is shown in Figure \ref{fig:system}: The illumination
system consists of four projectors that display various bright
patterns onto the three white walls in the corner of a dark
room.  The light reflects off of the walls and illuminates the
user's head indirectly.  After taking the frontal illuminations
we rotate the chair by 180 degrees and take pictures from the
opposite direction.  Having two cameras speeds the process
since only the chair needs to be moved in between frontal and
rear illuminations. Our projector-based system has several
advantages over flash-based illumination systems for face recognition:
\begin{itemize}
\item The illuminations can be modified in software, rather than hardware.
\item It is easy to capture many different illuminations quickly.
\item Good coverage and distant illumination can be achieved simultaneously.
\item There is no need to mount anything on the walls or construct a large dome.
\item The system can be assembled from off-the-shelf hardware.
\end{itemize}
\begin{figure}
\centering
\includegraphics[height=2in]{figures_cvpr/camera_rig.pdf}
\caption{{\bf Training acquisition system:} Four projectors and two cameras controlled by one computer.}
\vspace{-2em}
\label{fig:system}
\end{figure}
With our projector system, our choice of illuminations is
constrained only by the need to achieve a good
SNR\footnote{Since illuminations with more pixels illuminated
will have a better SNR (provided they don't saturate), there is
an engineering tradeoff between the SNR and the number of
training images.}, avoid saturation, and achieve
a reasonably short acquisition time.  Two simplifying
assumptions that we make are that every pixel is either turned
fully on or off in every illumination, and that the illuminated
regions do not overlap.

@@

\begin{figure}
\centering
\begin{tabular}{@{}c@{}c@{}}
\includegraphics[height=1.2in]{figures_cvpr/illum_results/coverage_sunset.pdf} &
\includegraphics[height=1.2in]{figures_cvpr/illum_results/granularity_sunset.pdf} \\
(a) Coverage & (b) Granularity
\end{tabular}\vspace{0mm}
\caption{{\bf Study of sufficient illuminations.} The average $\ell^1$ registration residual versus different illumination training sets. }
\label{fig:illumination-sufficiency}
\vspace{-4.5em}
\end{figure}

@@

See below for the 38 training images of one subject:
\begin{figure}[h]
\centering
\includegraphics[width=3.2in]{figures_cvpr/training.png}
\vspace{-2em}
\end{figure}

@@

\begin{table}
\caption{Recognition rates on the Multi-PIE database for
Algorithm 1 and \cite{Yang2010-CVPR}}
\centerline{
\begin{tabular}{|c|c|c|c|c| }
\hline
Recognition rate & Session 2 & Session 3 & Session 4  \\
\hline
{Alg. 1, $S=1$} & 90.7\% & 89.6\% & 87.5\% \\
\hline
{Alg. 1} & 93.9\% & 93.8\% & 92.3\% \\
\hline
{Alg. 1 with improved window} & 95.0\% & {\bf 96.3}\% & {\bf 97.3}\% \\
\hline
\cite{Yang2010-CVPR} & {\bf 95.2}\% & 93.4\% & 95.1\% \\
\hline
\end{tabular}
\label{tab:MultiPIE-recognition2} }
\end{table}

\newcommand{\tempwidth}[0]{0.58in}
\begin{figure}
\centering
{\small
\begin{tabular}{@{}c@{}c@{}c@{}c@{}c@{}c@{}}
\hspace{-2mm}\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/079_01_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/111_01_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/196_01_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/130_01_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/163_01_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/175_01_01_051_08.png} \\
\hspace{-2mm}\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/079_02_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/111_02_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/196_02_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/130_02_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/163_02_01_051_08.png}  &
\includegraphics[width=\tempwidth,clip=true]{figures_pami/multipie_failed/175_02_01_051_08.png} \\
\hspace{-2mm}(a) & (b) & (c) & (d) & (e) & (f) \vspace{0mm}
\end{tabular}
}
\caption{{\bf Representative failures from Multi-PIE}. {\bf Top:} training from Session 1; {\bf Bottom:} test images from Session 2. Due to changes in hair, glasses, beard, or pose, our alignment fails on these subjects regardless of test image illumination.}
\label{fig:failed-examples}
\vspace{-4mm}
\end{figure}
